Efficiently Handling Guards in the Low Level Design of RPython's Tracing JIT

### 2012-08-09 talk/vmil2012/paper.tex
a very preliminary attempt at a figure for vizualizing resume data
(cfbolz, bivab) generate and embed table about resume data sizes
Add a table showing the percentage of guards that ever fail and the percentage of guards that fail more than 200 times
remove some todos and update one
Use SSA reference

\subsection{RPython's Tracing JIT Compilers}
\label{sub:tracing}
Tracing JITs are a technique of just-in-time compilers that generate code by
observing the execution of a program. VMs using tracing JITs are typically
mixed mode execution environments containing also an interpreter. The
interpreter profiles the executed program and selects frequently executed code
paths to be compiled to machine code. After profiling identified an interesting
path, tracing is started, recording all operations that are executed on this
path. Like in most compilers tracing JITs use an intermediate representation
to store the recorded operations, which is typically in SSA form\todo{some ssa
reference}. Since tracing follows actual execution the code that is recorded
represents only one possible path through the control flow graph. Points of
divergence from the recorded path are marked with special operations called
\emph{guards}, these operations ensure that assumptions valid during the
tracing phase are still valid when the code has been compiled and is executed.
After a trace has been recorded it is optimized and then compiled to platform
specific machine code.

@@

\begin{itemize}
    \item For small integer constants
        the payload contains the value of the constant.
    \item For other constants
        the payload contains an index into a per-loop list of constants.
    \item For SSA variables,
        the payload is the number of the variable.
    \item For virtuals,
        the payload is an index into a list of virtuals, see next section.
\end{itemize}
\todo{figure showing linked resume-data}

@@

@@

\todo{figure about failure counts of guards (histogram?)}
\todo{add resume data sizes without sharing}
\todo{add a footnote about why guards have a threshold of 100}

>>>>

\subsection{RPython's Tracing JIT Compilers}
\label{sub:tracing}
Tracing JITs are a technique of just-in-time compilers that generate code by
observing the execution of a program. VMs using tracing JITs are typically
mixed mode execution environments containing also an interpreter. The
interpreter profiles the executed program and selects frequently executed code
paths to be compiled to machine code. After profiling identified an interesting
path, tracing is started, recording all operations that are executed on this
path. Like in most compilers tracing JITs use an intermediate representation to
store the recorded operations, which is typically in SSA
form~\cite{cytron_efficiently_1991}. Since tracing follows actual execution the
code that is recorded
represents only one possible path through the control flow graph. Points of
divergence from the recorded path are marked with special operations called
\emph{guards}, these operations ensure that assumptions valid during the
tracing phase are still valid when the code has been compiled and is executed.
After a trace has been recorded it is optimized and then compiled to platform
specific machine code.

@@

\begin{itemize}
    \item For small integer constants
        the payload contains the value of the constant.
    \item For other constants
        the payload contains an index into a per-loop list of constants.
    \item For SSA variables,
        the payload is the number of the variable.
    \item For virtuals,
        the payload is an index into a list of virtuals, see next section.
\end{itemize}

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{figures/resume_data.pdf}
\caption{The resume data for Figure~\ref{fig:trace-log}}
\label{fig:resume-data}
\end{figure}

@@

\begin{figure}
    \include{figures/resume_data_table}
    \caption{Resume Data sizes in KiB}
    \label{fig:resume_data_sizes}
\end{figure}

\begin{figure}
    \include{figures/failing_guards_table}
    \caption{Failing guards}
    \label{fig:failing_guards}
\end{figure}

\todo{add a footnote about why guards have a threshold of 200}
